Growth model analysis 3spp WB

### 2019-03-20 MS_Analysis_3spp_4rivers_Patterns_from 2spp.Rmd
add modCSR_add

##########################################################
# by cohortSpeciesRiver - model is overfit?

modCSR <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiverOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modCSR)
#plot(modCSR)

predGridCSR <- expand.grid(cohortSpeciesRiver = unique(modCSR$model$cohortSpeciesRiver),
                           ydayCumul = seq(250,1000, by = 30)) %>%
                    mutate(cohortOrdered = as.numeric(substr(cohortSpeciesRiver,1,4)),
                           speciesOrdered = substr(cohortSpeciesRiver,6,8),
                           riverOrdered = substring(cohortSpeciesRiver,10),
                           cohortCentered = cohortOrdered - cohortOffset,
                           species = speciesOrdered,
                           river = riverOrdered)

@@

ggplot(predGrid2CSR, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(50,210) +
  facet_grid(river ~ cohort)

##########################################################
# by speciesRiver ##########################################################

modSR <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modSR)
#plot(modSR, scale = 0)

predGridSR <- expand.grid(speciesRiverOrdered = unique(modSR$model$speciesRiverOrdered),
                          cohortOrdered = 2000:2013,
                          ydayCumul = seq(250,1000, by = 30)) %>%
                   mutate(speciesOrdered = substr(speciesRiverOrdered,1,3),
                          riverOrdered = substring(speciesRiverOrdered,5),
                          cohortCentered = cohortOrdered - cohortOffset,
                          species = speciesOrdered,
                          river = riverOrdered)

@@

# by speciesCohort
system.time(
modSC <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
)
summary(modSC)
#plot(modSC, scale = 0)

predGridSC <- expand.grid(speciesCohortOrdered = unique(modSC$model$speciesCohortOrdered),
                          river = unique(vB$riverOrdered),
                          ydayCumul = seq(250,1000, by = 30)) %>%
                   mutate(speciesOrdered = substr(speciesCohortOrdered,1,3),
                          cohortOrdered = as.numeric(substring(speciesCohortOrdered,5)),
                          cohortCentered = cohortOrdered - cohortOffset,
                          species = speciesOrdered)

@@

modCR <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered), 
               data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modCR)
#plot(modCR, scale = 0)

@@

# add year to smooths

modCSRY <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiverOrdered) +
               s(year, bs = 'cr', by = speciesOrdered), 
               data = vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt','bnt')), 
               method = "REML")
summary(modCSRY)

@@

mod <- bam(observedLength ~ cohortCentered * species * river + 
               s(ydayCumul, bs = 'cr'), 
               data = vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt','bnt')), 
               method = "REML")
summary(modCSRY)

AIC(modCSRY,modCSR,modSR,modSC,modCR,mod)  %>% rownames_to_column() %>% arrange(AIC)

>>>>

##########################################################
# by cohortSpeciesRiver

modCSR <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiverOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modCSR)
#plot(modCSR)

predGridCSR <- expand.grid(cohortSpeciesRiverOrdered = unique(modCSR$model$cohortSpeciesRiverOrdered),
                           ydayCumul = seq(250,1000, by = 30)) %>%
                    mutate(cohortOrdered = as.numeric(substr(cohortSpeciesRiverOrdered,1,4)),
                           speciesOrdered = substr(cohortSpeciesRiverOrdered,6,8),
                           riverOrdered = substring(cohortSpeciesRiverOrdered,10),
                           cohortRiverOrdered = paste(cohortOrdered, riverOrdered, sep = '_'),
                           speciesCohortOrdered = paste(speciesOrdered, cohortOrdered, sep = '_'),
                           speciesRiverOrdered = paste(speciesOrdered, riverOrdered, sep = '_'),
                           cohortCentered = cohortOrdered - cohortOffset,
                           cohort = cohortOrdered,
                           species = speciesOrdered,
                           river = riverOrdered)

@@

ggplot(predGrid2CSR, aes(ydayCumul,predCSR, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(50,210) +
  facet_grid(river ~ cohort)

##########################################################
# by cohortSpeciesRiver - additive splines
system.time(
modCSR_add <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
)
summary(modCSR_add)
#plot(modCSR_add)

predGridCSR_add <- expand.grid(cohortSpeciesRiverOrdered = unique(modCSR_add$model$cohortSpeciesRiverOrdered),
                           ydayCumul = seq(250,1000, by = 30)) %>%
                    mutate(cohortOrdered = as.numeric(substr(cohortSpeciesRiverOrdered,1,4)),
                           speciesOrdered = substr(cohortSpeciesRiverOrdered,6,8),
                           riverOrdered = substring(cohortSpeciesRiverOrdered,10),
                           cohortRiverOrdered = paste(cohortOrdered, riverOrdered, sep = '_'),
                           speciesCohortOrdered = paste(speciesOrdered, cohortOrdered, sep = '_'),
                           speciesRiverOrdered = paste(speciesOrdered, riverOrdered, sep = '_'),
                           cohortCentered = cohortOrdered - cohortOffset,
                           cohort = cohortOrdered,
                           species = speciesOrdered,
                           river = riverOrdered)

predCSR_add <- predict(modCSR_add, newdata = predGridCSR_add)#, type = 'lpmatrix')

predGrid2CSR_add <- bind_cols(predGridCSR_add,as.data.frame(predCSR_add))

ggplot(predGrid2CSR_add, aes(ydayCumul,predCSR_add, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(50,210) +
  facet_grid(river ~ cohort)

##########################################################
# by speciesRiver

modSR <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modSR)
#plot(modSR, scale = 0)

predGridSR <- expand.grid(speciesRiverOrdered = unique(modSR$model$speciesRiverOrdered),
                          cohortOrdered = 2000:2013,
                          ydayCumul = seq(250,1000, by = 30)) %>%
                   mutate(speciesOrdered = substr(speciesRiverOrdered,1,3),
                          riverOrdered = substring(speciesRiverOrdered,5),
                          cohortCentered = cohortOrdered - cohortOffset,
                          cohort = cohortOrdered,
                          species = speciesOrdered,
                          river = riverOrdered)

@@

# by speciesCohort
system.time(
modSC <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
)
summary(modSC)
#plot(modSC, scale = 0)

predGridSC <- expand.grid(speciesCohortOrdered = unique(modSC$model$speciesCohortOrdered),
                          river = unique(vB$riverOrdered),
                          ydayCumul = seq(250,1000, by = 30)) %>%
                   mutate(speciesOrdered = substr(speciesCohortOrdered,1,3),
                          cohortOrdered = as.numeric(substring(speciesCohortOrdered,5)),
                          cohortCentered = cohortOrdered - cohortOffset,
                          cohort = cohortOrdered,
                          species = speciesOrdered)

@@

modCR <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered), 
               data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(modCR)
#plot(modCR, scale = 0)

@@

# add year to smooths, model seems overfit

# try running without 'speciesOrdered' for year  this does a difference from the main fit, i.e. try s(year, bs = 'cr', by = speciesFactor)

modCSRY <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesOrdered) +
               s(ydayCumul, bs = 'cr', by = riverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesCohortOrdered) +
               s(ydayCumul, bs = 'cr', by = speciesRiverOrdered) +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiverOrdered) +
               #s(year, bs = 'cr', by = speciesOrdered), 
               s(year, bs = 'cr', by = speciesFactor),
               data = vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt','bnt')), 
               method = "REML")
summary(modCSRY)
#plot(modCSRY, select = 189)

predGridCSRY <- expand.grid(cohortSpeciesRiverOrdered = unique(modCSRY$model$cohortSpeciesRiverOrdered),
                            year = unique(modCSRY$model$year),
                            ydayCumul = seq(250,1000, by = 30)) %>%
                    mutate(cohortOrdered = as.numeric(substr(cohortSpeciesRiverOrdered,1,4)),
                           speciesOrdered = substr(cohortSpeciesRiverOrdered,6,8),
                           riverOrdered = substring(cohortSpeciesRiverOrdered,10),
                           cohortRiverOrdered = paste(cohortOrdered, riverOrdered, sep = '_'),
                           speciesCohortOrdered = paste(speciesOrdered, cohortOrdered, sep = '_'),
                           speciesRiverOrdered = paste(speciesOrdered, riverOrdered, sep = '_'),
                           cohortCentered = cohortOrdered - cohortOffset,
                           cohort = cohortOrdered,
                           species = speciesOrdered,
                           river = riverOrdered)

predCSRY <- predict(modCSRY, newdata = predGridCSRY)#, type = 'lpmatrix')

predGrid2CSRY <- bind_cols(predGridCSRY,as.data.frame(predCSRY))

ggplot(predGrid2CSRY, aes(ydayCumul, predCSRY, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(50,300) +
  facet_grid(river ~ cohortOrdered)

@@

mod <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr'), 
               data = vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt','bnt')), 
               method = "REML")
summary(modCSRY)

save(mod,modCR,modCSR,modSC,modSR,modCSRY, file = './data/out/mods.RData')
AIC(modCSR,modSR,modSC,modCR,mod)  %>% rownames_to_column() %>% arrange(AIC)
